Cluster Difference Imaging Photometric Survey. III. Rotation and Lithium in an Open Cluster Spanning 500\,Parsecs

### 2021-04-12 cluster_paper/ms.tex
update ms
update appendix 1

%
% NB. the "classical tidal radius" is from e.g., Piskunov+08 or
% Kharchenko+2013. The latter givs a radius r2 of 0.71 degrees. At 380
% pc, corresponds to 4.7pc. The King tidal radius given by K13 is
% 7.70pc.
%
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations surrounding nearby open clusters.  It is
  important to verify that these ``halos'', ``tails'', and ``strings''
  are of similar ages and compositions as stars in the denser part of
  the cluster.  We present an analysis of NGC\,2516
  ($\approx$150\,Myr), which has a classical tidal radius of 10\,pc
  and an apparent halo of stars spanning 500\,pc.  Combining
  photometry from Gaia, rotation periods from TESS, and lithium
  measurements from Gaia-ESO and GALAH, we find that the halo of
  NGC\,2516 is real and coeval with the cluster's core.  Two thirds of
  kinematically selected halo members out to 250\,pc from the cluster
  center have rotation periods consistent with a gyrochronological age
  of 150\,Myr.  A comparison sample of field stars shows no such
  trend.  In addition, the lithium abundances of stars in the halo are
  elevated relative to the field, and are correlated with the stellar
  rotation rate in a manner consistent with previous observations of
  similar clusters.  Broadly speaking, this work advances a paradigm
  in which the halos of open clusters are often more populous than
  their cores.  We highlight the implications for spectroscopic survey
  targeting, open cluster dispersal, and planet searches around young
  stars.
\end{abstract}

@@

One possible explanation is that the observed halo structures are
tidal tails \citep[{\it e.g.},][]{krumholz_star_2019}. The idea would
be that as the cluster contracts, stars escape out of the Lagrange
points imposed by the Milky Way's potential, and form leading and
lagging arms due to differential rotation in the Galaxy.  Kinematic
evidence that the core of \cn\ may be undergoing contraction has
recently been presented by \citet{healy_stellar_2020}, and might
support this interpretation.  Whether the exact rate of contraction,
and the corresponding ``evaporation rate'' of the core is sufficient
to explain the formation of the halo has yet to be assessed.

@@

\appendix
\section{Clustering methods and outcomes}
\label{app:clustering}

\begin{figure*}[h]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.4\textwidth]{f10.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Different clustering techniques yield different
  outcomes.} \nkinematic\ unique candidate cluster members found
  using three different techniques are considered in this study.
  \citetalias{cantatgaudin_gaia_2018}: \citet{cantatgaudin_gaia_2018},
  \citetalias{kounkel_untangling_2019}:
  \citet{kounkel_untangling_2019}, \citetalias{meingast_2021}:
  \citet{meingast_2021}.
  \label{fig:venn}
	}
\end{figure*}

Figure~\ref{fig:venn} is a Venn diagram showing the overlap between
the three membership catalogs concatenated in this study.  99\% of the
\citetalias{cantatgaudin_gaia_2018} sample overlaps with
\citetalias{kounkel_untangling_2019}, \citetalias{meingast_2021}, or
both.  By comparison, 36\% of the \citetalias{kounkel_untangling_2019}
sample, and 15\% of the \citetalias{meingast_2021} sample are
non-overlapping.
The data---Gaia DR2---used by all the studies was the same.
What are the differences in methodology that produce these different
outcomes?

%
% TODO: the following two paragraphs, WHEN QUOTED, are ENTIRELY FROM
% the relevant papers. they should probably be nixed in favor of
% concision.
%
\citetalias{cantatgaudin_gaia_2018} applied a procedure that 
yielded what we call ``the core''.  Their procedure was to first query
a Gaia DR2 cone around the previously reported RA and dec of the
cluster, and within $\pm 0.5$\,mas of its previously reported
distance. 
% TODO: which was it?
The outer radius of their cone was either r2 from MWSC (CITE
Kharchenko 2013), or twice the ``cluster radius'' listed by DAML/Dias (CITE).  No
proper motion cut was applied.  They then applied an unsupervised
classification scheme called UPMASK to $G<18$\,mag stars within this
cone (CITE).  The steps of UPMASK are first to perform a k-means
clustering in the ``astrometric space'' ($\mu_{\alpha'}, \mu_\delta,
\pi$).  Then, a ``veto'' step is applied to assess whether the groups
of stars output from the k-means clustering are or are not more
concentrated than a random distribution.  This is implemented by
``comparing the total branch length of the minimum spanning tree
connecting the stars with the expected branch length for a random
uniform distribution covering the investigated field of view''.  ``To
turn this yes/no flag to a membership probability,
\citet{cantatgaudin_gaia_2018} then redraw new values of
($\mu_{\alpha'}, \mu_\delta, \pi$) for each source based on the listed
value, uncertainty, and covariance.  After a certain number of
redrawings, the final probability is the frequency with which a given
star passes the veto".  In the case of \cn, this yielded a reported
``\texttt{r50}'' within which half of the cluster members were found
to be within 0.496$^\circ$.
When we selected candidate \cn\ members from the results of
\citetalias{cantatgaudin_gaia_2018}, we opted to include all candidate members
with reported membership probability exceeding 10\%.
While this seems {\it a priori} low, our results ({\bf SECTION XXX})
show that this ``membership probabilty'' severely underestimates the
purity of the \citetalias{cantatgaudin_gaia_2018} sample for \cn. 
%TODO: quantify
Their false positive rate across the sample is more like 1-5\%.

\citetalias{kounkel_untangling_2019}
applied a different unsupervised
clustering method to the 5-dimensional Gaia DR2 data (omitting radial
velocities, due to their sparsity).  Their selection function (see
their Section~2) yielded $\approx 2\times 10^7$ stars, mostly within
$\approx 1$\,kpc and typically with $G<18$\,mag.  Their clustering
algorithm, which was run on this entire  stellar sample,
 was the ``hierarchical density-based spatial clustering of
applications with noise'' (HDBSCAN, CITE McInnes17).  The classical
DBSCAN algorithm ``identifies clusters as overdensities in a
multi-dimensional space in which the number of sources exceeds the
required minimum number of points within a neighborhood of a
particular linking length $\epsilon$.  HDBSCAN does not depend on
$\epsilon$; instead it condenses the minimum spanning tree by pruning
off the nodes that do not meet the minimum number of sources in a
cluster and reanalyzing the nodes that do. Depending on the chosen
algorithm, it would then either find the most persistent structure
(through the excess of mass method), or return clusters as the leaves
of the tree (which results in somewhat more homogeneous clusters). In
both cases it is more effective at finding structures of varying
densities in a given data set than DBSCAN.''
`` The two main parameters that control HDBSCAN are the number of
sources in a cluster and the number of samples. The former is the
parameter that rejects groupings that are too small; the latter sets
the threshold of how conservative the algorithm is in its
considerations of the background noise (even if the resulting noisy
groupings do meet the minimum cluster size).  By default, the sample
size is set to the same value as the cluster size, but it is possible
to adjust them separately.''
Regarding membership proabilities,
\citetalias{kounkel_untangling_2019} did not report continuous membership
probabilities, instead opting for the binary ``member'' or ``not''.

\citetalias{meingast_2021} did the same as
\citetalias{kounkel_untangling_2019}, but required A) the tangential
velocity dispersions to be smaller, and B) deconvolved the distances.
{\bf FIXME FIXME FIXME}
{\bf FIXME FIXME FIXME}
{\bf FIXME FIXME FIXME}

\citet{meingast_2021} reported a 
Finally, 1577 of the 1860 \citet{meingast_2021} sources (85\%)
were included in the \citet{kounkel_untangling_2019} sample.

>>>>

%
% NB. the "classical tidal radius" is from e.g., Piskunov+08 or
% Kharchenko+2013. The latter givs a radius r2 of 0.71 degrees. At 380
% pc, corresponds to 4.7pc. The King tidal radius given by K13 is
% 7.70pc.
%
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations surrounding nearby open clusters.  It is
  important to verify that these ``halos'', ``tails'', and ``strings''
  are of similar ages and compositions as stars in the denser part of
  the cluster.  We present an analysis of NGC\,2516
  ($\approx$150\,Myr), which has a classical tidal radius of 10\,pc
  and an apparent halo of stars spanning 500\,pc.  Combining
  photometry from Gaia, rotation periods from TESS, and lithium
  measurements from Gaia-ESO and GALAH, we find that the halo of
  NGC\,2516 is real and coeval with the cluster's core.  Two thirds of
  kinematically selected halo members out to 250\,pc from the cluster
  center have rotation periods consistent with a gyrochronological age
  of 150\,Myr.  A comparison sample of field stars shows no such
  trend.  In addition, the lithium abundances of stars in the halo are
  elevated relative to the field, and are correlated with the stellar
  rotation rate in a manner consistent with previous observations of
  similar clusters.  Broadly speaking, this work advances a paradigm
  in which the halos of open clusters are often more populous than
  their cores.  There are significant implications for spectroscopic
  survey targeting, open cluster dispersal, and planet searches around
  young stars.
\end{abstract}

@@

One possible explanation is that the observed halo structures are
tidal tails \citep[{\it e.g.},][]{krumholz_star_2019}. The idea would
be that as the cluster contracts, stars escape out of the Lagrange
points imposed by the Milky Way's potential, and form leading and
lagging arms due to differential rotation in the Galaxy.  Kinematic
evidence that the core of \cn\ may be undergoing contraction has
recently been presented by \citet{healy_stellar_2020}, and might
support this interpretation.  Whether the exact rate of contraction,
and the corresponding ``evaporation rate'' of the core is sufficient
to explain the formation of the halo has yet to be assessed.  The
presence or lack of epicyclic overdensities in the tails might also
help clarify whether this scenario is happening
\citep{kupper_more_2012}.

@@

\appendix
\section{Clustering Methods and Outcomes}
\label{app:clustering}

\begin{figure}[h]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.3\textwidth]{f10.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Different clustering techniques yield different
  candidate members of \cn.} \nkinematic\ unique candidate cluster
  members found using three different techniques are considered in
  this study.  \citetalias{cantatgaudin_gaia_2018}:
  \citet{cantatgaudin_gaia_2018},
  \citetalias{kounkel_untangling_2019}:
  \citet{kounkel_untangling_2019}, \citetalias{meingast_2021}:
  \citet{meingast_2021}.
  \label{fig:venn}
	}
\end{figure}

Figure~\ref{fig:venn} is a Venn diagram of
the three membership catalogs concatenated in this study.  99\% of the
\citetalias{cantatgaudin_gaia_2018} sample overlaps with
\citetalias{kounkel_untangling_2019}, \citetalias{meingast_2021}, or
both.  By comparison, 36\% of the \citetalias{kounkel_untangling_2019}
sample, and 15\% of the \citetalias{meingast_2021} sample do not
overlap with any of the other catalogs.  The data---Gaia DR2---used by
all the studies was the same.  What are the differences in methodology
that produce these different outcomes?

\citetalias{cantatgaudin_gaia_2018} applied a procedure that yielded
what we colloquially call ``the core''.  Their membership assignment
algorithm was to first query a Gaia DR2 cone around the previously
reported $\{\alpha,\delta\}$ of the cluster center, and within $\pm
0.5$\,mas of its previously reported parallax.  The outer radius of
their cone was the previously reported angular radius of the cluster
(0.71$^\circ$; \citealt{Kharchenko_et_al_2013}).  No proper motion cut
was applied.  \citetalias{cantatgaudin_gaia_2018} then applied an
unsupervised classification scheme to $G<18$\,mag stars within this
cone (UPMASK; \citealt{kronemartins_upmask_2014}).  The UPMASK
algorithm first performs a $k$-means clustering on the astrometric
parameters, $\{\mu_{\alpha'}, \mu_\delta, \pi\}$.  A ``veto''
step is then applied to assess whether the groups of stars output from the
$k$-means clustering are more concentrated than a uniform
distribution, by comparing the total branch lengths of their
respective minimum spanning trees.  This binary flag is converted to a
membership probability by redrawing values for $\{\mu_{\alpha'},
\mu_\delta, \pi\}$ using the reported uncertainties and
covariances.  The final membership probability for any given star is
then the fraction of draws for which the star passes the veto step.
In the case of \cn, the resulting radius within which half of the
cluster members were found was 0.50$^\circ$.  We included all
\citetalias{cantatgaudin_gaia_2018} \cn\ members with reported
membership probability exceeding 10\% in our list of candidate cluster
members.

\citetalias{kounkel_untangling_2019} applied a different unsupervised
clustering method to the 5-D Gaia DR2 positions and proper motions.
\citetalias{kounkel_untangling_2019} began with a sample of
$\approx$$2\times 10^7$ stars, mostly within $\approx 1$\,kpc and with
$G\lesssim18$\,mag.  They then applied the hierarchical density-based
spatial clustering of applications with noise algorithm (HDBSCAN;
\citealt{campello_hierarchical_2015}, \citealt{mcinnes_hdbscan_2017})
on the entire sample,  setting the minimum number of stars per cluster
to be 40.  \citetalias{kounkel_untangling_2019} then iterated over a
few different cutoff parallaxes to improve sensitivity to structures
with distances from the Sun ranging between $\approx$100\,pc and
$\approx$1000\,pc.  They then fitted ages to the resulting clusters
using two different methods: isochronal, and a convolutional neural
network.  Regarding membership proabilities,
\citetalias{kounkel_untangling_2019} reported the binary ``member'' or
``not'' from HDBSCAN.
We included all \citetalias{kounkel_untangling_2019} \cn\ members in
our list of candidate cluster members.

Finally, \citetalias{meingast_2021} considered ten known open clusters
within the nearest kiloparsec. One of these clusters was \cn.  Their
initial  list of $\approx$$3\times10^7$ was similar to that of
\citetalias{kounkel_untangling_2019}, mainly requiring well-behaved
astrometry and photometry from Gaia.  Given the mean positions and
velocities of the clusters determined by
\citetalias{cantatgaudin_gaia_2018}, \citetalias{meingast_2021} then
selected all stars with proper-motion correct tangential velocities
within 1.5\,\kms\ of the cluster mean.  They then applied DBSCAN
\citep{ester_density-based_1996} on both the observed galactic
$\{X,Y,Z\}$ coordinates, as well as to a separate deconvolved spatial
coordinate distribution discussed in their Section~3.3.  The results
of their procedure, as well as those of
\citetalias{cantatgaudin_gaia_2018} and
\citetalias{kounkel_untangling_2019}, are visible at the website of
S.~Meingast:
\url{https://homepage.univie.ac.at/stefan.meingast/coronae.html}.
Their reported membership proabilities were binary, similar to those
of \citetalias{kounkel_untangling_2019}.  All of the
\citetalias{meingast_2021} \cn\ members were included in our list of
candidate cluster members.
